---
title: "Automatic Differentiation with dualr"
author: "Alexander Towell"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Automatic Differentiation with dualr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
has_dualr <- requireNamespace("dualr", quietly = TRUE)
```

## Why Automatic Differentiation?

Maximum likelihood estimation requires derivatives of the log-likelihood:
the **score** (gradient) for first-order methods and the **Hessian** (or
observed information) for second-order methods like Newton-Raphson. There
are three ways to supply these:

| Approach | Accuracy | Effort | Speed |
|----------|----------|--------|-------|
| Numerical (`numDeriv`) | $O(h^2)$ truncation error | None (default) | Slow (extra evaluations) |
| Hand-coded analytic | Exact | High (error-prone) | Fast |
| Automatic differentiation (`dualr`) | Exact (to machine precision) | Low | Moderate |

Following the SICP principle that *the specification should be separate
from the mechanism*, the log-likelihood function is the specification of
our statistical model. Derivatives are mechanical consequences of that
specification---they should be derived automatically rather than
hand-coded.

The [`dualr`](https://github.com/queelius/dualr) package provides
forward-mode automatic differentiation for R via dual numbers, giving
exact derivatives with minimal code changes.

## Setup

```{r load-packages, eval=has_dualr}
library(compositional.mle)
library(dualr)
```

## The Three Approaches

Consider a Poisson model. The log-likelihood for $\lambda$ given data
$x_1, \ldots, x_n$ is:

$$\ell(\lambda) = \left(\sum_{i=1}^n x_i\right) \log \lambda - n\lambda$$

```{r poisson-data, eval=has_dualr}
set.seed(42)
x <- rpois(100, lambda = 3.5)
```

We define the log-likelihood once:

```{r poisson-loglik, eval=has_dualr}
ll_poisson <- function(theta) {
  lambda <- theta[1]
  n <- length(x)
  # Accumulate sum in a loop for dualr compatibility
  sx <- 0
  for (i in seq_along(x)) sx <- sx + x[i]
  sx * log(lambda) - n * lambda
}
```

Now we create three problem specifications using different derivative
strategies:

```{r three-approaches, eval=has_dualr}
# 1. No derivatives — numDeriv fallback (the default)
p_numerical <- mle_problem(loglike = ll_poisson)

# 2. Hand-coded analytical derivatives
p_analytic <- mle_problem(
  loglike = ll_poisson,
  score = function(theta) sum(x) / theta[1] - length(x),
  fisher = function(theta) matrix(sum(x) / theta[1]^2, 1, 1)
)

# 3. Automatic differentiation via dualr
p_ad <- mle_problem(
  loglike = ll_poisson,
  score = function(theta) dualr::score(ll_poisson, theta),
  fisher = function(theta) dualr::observed_information(ll_poisson, theta)
)
```

```{r three-approaches-print, eval=has_dualr}
p_numerical
p_ad
```

Notice that both the AD and analytic problems report "analytic" score and
Fisher information. From the solver's perspective, they are identical---the
solver doesn't know (or care) whether the derivatives were hand-coded or
computed by AD. This is the power of the `mle_problem` abstraction:
derivatives are pluggable.

## Comparison

Let's run the same solver on all three:

```{r comparison, eval=has_dualr}
solver <- bfgs(max_iter = 200)

r_num <- solver(p_numerical, theta0 = 1)
r_ana <- solver(p_analytic, theta0 = 1)
r_ad  <- solver(p_ad, theta0 = 1)

results <- data.frame(
  Method     = c("numDeriv", "Analytic", "dualr AD"),
  Estimate   = c(r_num$theta.hat, r_ana$theta.hat, r_ad$theta.hat),
  LogLik     = c(r_num$loglike, r_ana$loglike, r_ad$loglike),
  Converged  = c(r_num$converged, r_ana$converged, r_ad$converged),
  Iterations = c(r_num$iterations, r_ana$iterations, r_ad$iterations)
)
results
```

```{r true-mle, eval=has_dualr}
cat("True MLE (sample mean):", mean(x), "\n")
```

All three converge to the same estimate. The key differences are:

- **numDeriv** introduces small $O(h^2)$ errors in the gradient, which
  may cause minor differences in iteration paths (usually negligible)
- **Analytic** and **dualr AD** provide exact gradients, leading to
  identical optimization paths
- **dualr AD** requires no manual derivative derivation

## Multivariate Example: Normal Distribution

For the Normal$(\mu, \sigma)$ model, we have two parameters and a
positivity constraint on $\sigma$. The log-likelihood is:

$$\ell(\mu, \sigma) = -n \log \sigma - \frac{1}{2\sigma^2}\sum_{i=1}^n (y_i - \mu)^2$$

```{r normal-setup, eval=has_dualr}
set.seed(123)
y <- rnorm(200, mean = 5, sd = 2)
```

```{r normal-loglik, eval=has_dualr}
# Log-likelihood with loop-based sum for dualr Hessian support
ll_normal <- function(theta) {
  mu <- theta[1]
  sigma <- theta[2]
  n <- length(y)
  ss <- 0
  for (i in seq_along(y)) {
    ss <- ss + (y[i] - mu)^2
  }
  -n * log(sigma) - 0.5 * ss / sigma^2
}
```

**Note:** We use a loop-based sum instead of vectorized `sum((y - mu)^2)`.
This is because `dualr`'s Hessian computation currently requires scalar
accumulation when mixing dual numbers with data vectors. The score function
works either way, but the Hessian needs this pattern. This is a minor
syntactic cost for exact second derivatives.

```{r normal-ad, eval=has_dualr}
# AD-powered problem with box constraints via L-BFGS-B
p_normal <- mle_problem(
  loglike = ll_normal,
  score = function(theta) dualr::score(ll_normal, theta),
  fisher = function(theta) dualr::observed_information(ll_normal, theta)
)

solver <- lbfgsb(lower = c(-Inf, 1e-4), upper = c(Inf, Inf))
result <- solver(p_normal, theta0 = c(0, 1))
```

```{r normal-results, eval=has_dualr}
cat("Estimated mu:   ", round(result$theta.hat[1], 4), "\n")
cat("Estimated sigma:", round(result$theta.hat[2], 4), "\n")
cat("Converged:      ", result$converged, "\n")

# Compare with true MLE
cat("\nTrue MLE mu:   ", round(mean(y), 4), "\n")
cat("True MLE sigma:", round(sd(y) * sqrt((length(y)-1)/length(y)), 4), "\n")
```

The Hessian matrix from dualr provides the full curvature information:

```{r normal-hessian, eval=has_dualr}
H <- dualr::hessian(ll_normal, result$theta.hat)
cat("Hessian at MLE:\n")
print(round(H, 2))

cat("\nObserved information at MLE:\n")
I_obs <- dualr::observed_information(ll_normal, result$theta.hat)
print(round(I_obs, 2))

cat("\nApproximate standard errors (from observed information):\n")
se <- sqrt(diag(solve(I_obs)))
cat("  SE(mu):   ", round(se[1], 4), "\n")
cat("  SE(sigma):", round(se[2], 4), "\n")
```

## Composition with AD

The real power emerges when combining AD with solver composition. Because
`dualr`-derived derivatives are just functions, they compose seamlessly:

```{r composition, eval=has_dualr}
# Coarse grid search, then refine with L-BFGS-B
strategy <- grid_search(
  lower = c(0, 0.5), upper = c(10, 5), n = 5
) %>>% lbfgsb(lower = c(-Inf, 1e-4), upper = c(Inf, Inf))

result_composed <- strategy(p_normal, theta0 = c(0, 1))
cat("Grid -> L-BFGS-B estimate:",
    round(result_composed$theta.hat, 4), "\n")
cat("Converged:", result_composed$converged, "\n")
```

## When to Use What

| Situation | Recommended | Reason |
|-----------|-------------|--------|
| Quick prototyping | `numDeriv` (default) | Zero effort, just write the log-likelihood |
| Production / accuracy-critical | `dualr` | Exact derivatives, minimal code overhead |
| Maximum performance on simple models | Hand-coded | Avoids AD overhead for trivial derivatives |
| Complex models (mixtures, hierarchical) | `dualr` | Hand-coding is error-prone and tedious |
| Derivative-free problems | `nelder_mead()` | When the log-likelihood isn't differentiable |

The general recommendation is to **start with the default** (no explicit
derivatives) during model development, then **switch to `dualr`** when
you need accuracy or performance for production use. Hand-coding is
only worthwhile for very simple models where the derivatives are obvious.
